---
title: "Autistic vs Non-autistic VSOA"
author: "Chris Cox"
format: revealjs
editor: visual
---

# Setup

## Required Packages
```{r packages}
#| echo: true
#| cache: false

library(dplyr)
library(purrr)
library(tidyr)
library(knitr)
library(kableExtra)
library(ggplot2)
```

## Load metadata
```{r load metadata}
#| echo: true
#| cache: true

new_guids <- readRDS("data/autistic-guis-added-20250507.rds")
d_meta <- readRDS("data/cdi-metadata.rds")
d_cdi <- readRDS("data/asd_na-osg-2025-05-20.rds") |>
    mutate(
        dataset = if_else(subjectkey %in% new_guids, 2, 1),
        dataset = factor(dataset, 1:2, c("old", "new"))
    ) |>
    left_join(
        d_meta |> select(num_item_id, word)
    )
```

## Load VSOA
```{r load vsoa}
#| echo: true
#| cache: true

d_vsoa <- readRDS("data/vsoa-autistic-nonautistic-ndar-id-fix-remodel-v2.rds")
d_vsoa_diff <- readRDS("data/vsoa-autistic-nonautistic-diff-ndar-id-fix-remodel-v2.rds")
d_vsoa_diff_old <- readRDS("data/vsoa-autistic-nonautistic-diff-old.rds") |>
    rename(vsoa_NA = na, vsoa_ASD = asd, vsoa_diff = diff,
           ci_l_diff = ci_l, ci_u_diff = ci_u)
```

## Trim VSOA
```{r trim vsoa}
#| echo: true
#| cache: true

trim_vsoa <- function(.data) {
    .data |>
        mutate(
            vsoa_NA = if_else(vsoa_NA <   1,  1, vsoa_NA),
            vsoa_NA = if_else(vsoa_NA > 680, NA, vsoa_NA),
            vsoa_ASD = if_else(vsoa_ASD <   1,  1, vsoa_ASD),
            vsoa_ASD = if_else(vsoa_ASD > 680, NA, vsoa_ASD),
            vsoa_diff = vsoa_ASD - vsoa_NA
        ) |>
        drop_na()
}
d_vsoa_diff_trimmed <- trim_vsoa(d_vsoa_diff)
d_vsoa_diff_old_trimmed <- trim_vsoa(d_vsoa_diff_old)
```

## Prepare participant characteristics
```{r participant data}
#| echo: true
#| cache: true

d_ppt <- d_cdi |>
    select(subjectkey, nproduced, sex, group, interview_age) |>
    distinct()
```


## Load and merge AoA

```{r load merge aoa}
#| echo: true
#| cache: true

d_aoa <- readRDS("data/CDI_English (American)_WS_14May2025_AoA.rds")
d_vsoa_aoa <- left_join(
    d_vsoa |> filter(group != "ASD-NA"),
    d_aoa |> select(num_item_id, aoa = aoa_glm), by = "num_item_id"
) |>
    mutate(
        group = factor(group, c("ASD", "NA"), c("autistic", "not aut."))
    ) |>
    relocate(aoa, .before = vsoa)
```

## Trim AoA (and paired VSOA)

```{r trim aoa}
#| echo: true
#| cache: true

d_vsoa_aoa_trimmed <- d_vsoa_aoa |>
    mutate(
        vsoa = if_else(vsoa <   1,  1, vsoa),
        vsoa = if_else(vsoa > 680, NA, vsoa),
        aoa  = if_else( aoa <  12, 12,  aoa),
        aoa  = if_else( aoa >  30, NA,  aoa)
    ) |>
    drop_na()
```


# Participant Characteristics

## Vocabulary size and interview age {.smaller}
```{r vocab size interview age}
#| echo: false
#| cache: true

d_ppt |>
    mutate(
        group = factor(group, c("ASD", "NA"), c("autistic", "not aut."))
    ) |>
    group_by(group) |>
    summarize(
        across(
            c(nproduced, interview_age),
            list(mean = mean, median = median, sd = sd)
        )
    ) |> 
    kable(col.names = c("group", "mean", "median", "SD", "mean", "median", "SD"), digits = 3) |>
    kableExtra::add_header_above(header = c(" " = 1, "Productive Vocabulary" = 3, "Interview Age" = 3))
```

## Gender and Group

```{r gender group}
#| echo: false
#| cache: true

d_ppt |>
    mutate(
        group = factor(group, c("ASD", "NA"), c("autistic", "not aut.")),
        sex = factor(sex, c("M", "F"), c("male", "female"))
    ) |>
    count(group, sex) |>
    pivot_wider(id_cols = group, names_from = sex, values_from = n) |>
    mutate(total = male + female, `ratio (M:F)` = male / female) |>
    kable(digits = 3)
```


# Correlation Analysis
## Helper functions
```{r vsoa corr functions}
#| echo: true
#| cache: true

vsoa_corr_tbl <- function(.data, .x, .y, ..., digits = 3) {
    cor_methods <- c("pearson", "spearman", "kendall")
    map(cor_methods,
        function(df, method) {
            df |>
                ungroup() |>
                group_by(...) |>
                summarize(r = cor({{ .x }}, {{ .y }}, method = method,
                                  use = "complete.obs"))
        },
        df = .data
    ) |>
        set_names(cor_methods) |>
        list_rbind(names_to = "method") |>
        pivot_wider(names_from = method, values_from = r) |>
        kable(digits = digits)
}

vsoa_corr_fig <- function(.data, .x, .y, ...) {
    p <- ggplot(.data, aes(x = {{ .x }}, y = {{ .y }})) +
        geom_point() +
        geom_smooth(method = lm) +
        theme_bw(base_size = 18) +
        facet_wrap(vars(...))
    
    return(p)
}
```

## VSOA and AoA Tbl. (raw)

### Number of words contributing to correlations
```{r number of words in correlation analysis}
#| cache: true
d_vsoa_aoa |>
    drop_na() |>
    count(group) |>
    kable()
```

### Correlations
```{r vsoa corr metrics tbl}
#| echo: false
#| cache: true
vsoa_corr_tbl(d_vsoa_aoa, aoa, vsoa, group)
```

## VSOA and AoA Fig. (raw) {.smaller}
```{r vsoa corr metrics fig}
#| echo: false
#| cache: true
vsoa_corr_fig(d_vsoa_aoa, aoa, vsoa, group)
```


## VSOA and AoA Tbl. (trimmed)  {.smaller}

### Number of words contributing to correlations
```{r number of words in correlation analysis trimed}
d_vsoa_aoa_trimmed |>
    drop_na() |>
    count(group) |>
    kable()
```

### Correlations
```{r vsoa corr metrics trimmed tbl}
#| echo: false
#| cache: true
vsoa_corr_tbl(d_vsoa_aoa_trimmed, aoa, vsoa, group)
```

## VSOA and AoA Fig. (trimmed)
```{r vsoa corr metrics trimmed fig}
#| echo: false
#| cache: true
vsoa_corr_fig(d_vsoa_aoa_trimmed, aoa, vsoa, group)
```

## Autistic and non-autistic Tbl. (raw) {.smaller}

### Number of words contributing to correlations
```{r number of words in correlation aut non all}
d_vsoa_aoa_trimmed |>
    drop_na() |>
    count(group) |>
    kable()
```

### Correlations
```{r vsoa correlation tbl aut non all}
vsoa_corr_tbl(d_vsoa_diff, vsoa_ASD, vsoa_NA)
```


## Autistic and non-autistic Fig. (raw)
```{r vsoa correlation fig aut non all}
vsoa_corr_fig(d_vsoa_diff, vsoa_ASD, vsoa_NA)
```

## Autistic and non-autistic Tbl. (trimmed)  {.smaller}

```{r number of words in correlation aut non all trimed}
d_vsoa_aoa_trimmed |>
    drop_na() |>
    count() |>
    kable()
```

```{r vsoa correlation tbl aut non all trimmed}
vsoa_corr_tbl(d_vsoa_diff_trimmed, vsoa_ASD, vsoa_NA)
```


## Autistic and non-autistic Fig. (trimmed)  {.smaller}
```{r vsoa correlation fig aut non all trimmed}
vsoa_corr_fig(d_vsoa_diff_trimmed, vsoa_ASD, vsoa_NA)
```


# Inspecting VSOA estimates

## How many words have extreme VSOA?

* Low: $\textrm{VSOA} \le 1$
* High: $\textrm{VSOA} \ge \max \textrm{vocab size}$

```{r extreme vsoa}
#| echo: true
#| cache: true
#| output-location: slide

d_vsoa |>
    filter(group != "ASD-NA") |>
    mutate(
        extreme = if_else(vsoa <= 1, 1,
                          if_else(vsoa >= max(d_ppt$nproduced), 2, NA)),
        extreme = factor(extreme, 1:2, c("low", "high"))
    ) |>
    drop_na(extreme) |>
    count(group, extreme) |>
    pivot_wider(id_cols = group, names_from = extreme,
                values_from = n, values_fill = 0) |>
    kable()
```


## Least-produced words (for autistic only; counts) {.scrollable}

```{r least produced words}
d_cdi_cumulative <- bind_rows(
    old = d_cdi |> filter(dataset == "old"),
    new = d_cdi,
    .id = "dataset_cumulative"
)

least_produced <- d_cdi_cumulative |>
    group_by(dataset_cumulative, group, word) |>
    summarize(n = sum(produced)) |>
    group_by(dataset_cumulative, group) |>
    slice_min(n, n=10)
    
least_produced_x <- d_cdi_cumulative |>
    filter(word %in% least_produced$word) |>
    group_by(dataset_cumulative, group, word) |>
    summarize(n = sum(produced), p = n / n()) |>
    mutate(
        least_10_asd = word %in%
            pull(filter(least_produced, group == "ASD", dataset_cumulative == "old"), word),
        least_10_non = word %in%
            pull(filter(least_produced, group == "NA", dataset_cumulative == "old"), word),
        least_10_for = if_else(
            least_10_asd & least_10_non,
            "both",
            if_else(
                least_10_asd & !least_10_non,
                "ASD",
                "NA"
            )
        )
    )

least_produced_counts <- least_produced_x |>
    pivot_wider(
        id_cols = c(least_10_for, word),
        names_from = c(dataset_cumulative, group),
        values_from = n,
        values_fill = 0
    ) |>
    relocate(least_10_for, new_ASD, old_ASD, new_NA, old_NA, .after = word)

least_produced_proportions <- least_produced_x |>
    pivot_wider(
        id_cols = c(least_10_for, word),
        names_from = c(dataset_cumulative, group),
        values_from = p,
        values_fill = 0
    ) |>
    relocate(least_10_for, new_ASD, old_ASD, new_NA, old_NA, .after = word)
```


```{r least produced words for ASD only counts}
least_produced_counts |>
    filter(least_10_for == "ASD") |>
    select(-least_10_for) |>
    kable(col.names = c("word", "New", "Old", "New", "Old"), digits = 2) |>
    kableExtra::add_header_above(header = c(" " = 1, "Autistic" = 2, "Not Autistic" = 2))
```


## Least-produced words (for autistic only; proportions) {.scrollable}

```{r least produced words p}
least_produced_proportions |>
    filter(least_10_for == "ASD") |>
    select(-least_10_for) |>
    relocate(new_ASD, old_ASD, new_NA, old_NA, .after = word) |>
    kable(col.names = c("word", "New", "Old", "New", "Old"), digits = 2) |>
    kableExtra::add_header_above(header = c(" " = 1, "Autistic" = 2, "Not Autistic" = 2))
```



## Least-produced words (for non-autistic only; counts) {.scrollable}

```{r least produced words for NA only counts}
least_produced_counts |>
    filter(least_10_for == "NA") |>
    select(-least_10_for) |>
    kable(col.names = c("word", "New", "Old", "New", "Old"), digits = 2) |>
    kableExtra::add_header_above(header = c(" " = 1, "Autistic" = 2, "Not Autistic" = 2))
```


## Least-produced words (for non-autistic only; proportions) {.scrollable}

```{r least produced words for NA only proportions}
least_produced_proportions |>
    filter(least_10_for == "NA") |>
    select(-least_10_for) |>
    relocate(new_ASD, old_ASD, new_NA, old_NA, .after = word) |>
    kable(col.names = c("word", "New", "Old", "New", "Old"), digits = 2) |>
    kableExtra::add_header_above(header = c(" " = 1, "Autistic" = 2, "Not Autistic" = 2))
```


## Words with largest VSOA differences  {.smaller}

### Autistic children learn earlier
```{r autistic children learn earlier}
d_vsoa_diff |>
    select(-starts_with("ci_")) |>
    slice_min(vsoa_diff, n = 7) |>
    kable(digits = 1)
```



## Words with largest VSOA differences  {.smaller}

### Autistic children learn later
```{r autistic children learn later}
d_vsoa_diff |>
    select(-starts_with("ci_")) |>
    slice_max(vsoa_diff, n = 7) |>
    kable(digits = 1)
```


## Words with largest VSOA differences (trimmed)  {.smaller}

### Autistic children learn earlier
```{r autistic children learn earlier trimmed}
d_vsoa_diff_trimmed |>
    select(-starts_with("ci_")) |>
    slice_min(vsoa_diff, n = 7) |>
    kable(digits = 0)
```


## Words with largest VSOA differences (trimmed) {.smaller}

### Autistic children learn later
```{r autistic children learn later trimmed}
d_vsoa_diff_trimmed |>
    select(-starts_with("ci_")) |>
    slice_max(vsoa_diff, n = 7) |>
    kable(digits = 0)
```

# Significantly different VSOAs by group

## By lexical class {.smaller}

```{r sig diff tbl function}
sig_diff_tbl <- function(.data, .meta, .by) {
    by_name <- rlang::quo_name(.by)
    levels <- switch(by_name,
           lexical_class = c("adjectives", "function_words", "nouns", "verbs", "other", "total"),
           category = c()
    )
    tmp <- .data |>
        left_join(select(.meta, num_item_id, {{ .by }}))
    
    bind_rows(
        A = tmp,
        B = tmp |> mutate({{ .by }} := "total")
    ) |>
        mutate(
            {{ .by }} := factor(
                {{ .by }},
                levels = c("adjectives", "function_words",
                           "nouns", "verbs", "other", "total")),
            produced_earlier_by = if_else(sign(vsoa_diff) < 0, 1, 2) |>
                factor(1:2, c("autistic", "not aut.")),
            significant = sign(ci_u_diff) == sign(ci_l_diff)
        ) |>
        filter(significant) |>
        count(produced_earlier_by, {{ .by }}, .drop = FALSE) |>
        pivot_wider(
            id_cols = {{ .by }},
            names_from = produced_earlier_by,
            values_from = n,
            values_fill = 0
        ) |>
        mutate(ratio = autistic / `not aut.`) |>
        kable(digits = 2) |>
        kableExtra::add_header_above(
            header = c(" " = 1, "Produced earlier by" = 2, " " = 1)
        )
}
```


:::{.columns}

::: {.column width="50%"}
### New VSOAs
```{r significant differences lexical class new}
sig_diff_tbl(d_vsoa_diff, d_meta, lexical_class)
```
:::

::: {.column width="50%"}
### Old VSOAs
```{r significant differences lexical class old}
sig_diff_tbl(d_vsoa_diff_old, d_meta, lexical_class)
```
:::

:::

## By category {.smaller}

:::{.columns}

::: {.column width="50%"}
### New VSOAs
```{r significant differences category new}
sig_diff_tbl(d_vsoa_diff, d_meta, category)
```
:::

::: {.column width="50%"}
### Old VSOAs
```{r significant differences category old}
sig_diff_tbl(d_vsoa_diff_old, d_meta, category)
```
:::

:::


