---
title: "Autistic vs Non-autistic VSOA"
author: "Chris Cox"
format: revealjs
editor: visual
---

## Load data

```{r load data}
library(dplyr)
library(purrr)
library(tidyr)
library(knitr)
library(kableExtra)
library(ggplot2)

new_guids <- readRDS("data/autistic-guis-added-20250507.rds")
d_meta <- readRDS("data/cdi-metadata.rds")
d_cdi <- readRDS("data/asd_na-osg-2025-05-07.rds") |>
    mutate(
        dataset = if_else(subjectkey %in% new_guids, 2, 1),
        dataset = factor(dataset, 1:2, c("old", "new"))
    ) |>
    left_join(
        d_meta |> select(num_item_id, word)
    )
d_vsoa <- readRDS("data/vsoa-autistic-nonautistic.rds")
d_vsoa_diff <- readRDS("data/vsoa-autistic-nonautistic-diff.rds")
d_vsoa_diff_trimmed <- d_vsoa_diff |>
    mutate(
        vsoa_NA = if_else(vsoa_NA <   1,  1, vsoa_NA),
        vsoa_NA = if_else(vsoa_NA > 680, NA, vsoa_NA),
        vsoa_ASD = if_else(vsoa_ASD <   1,  1, vsoa_ASD),
        vsoa_ASD = if_else(vsoa_ASD > 680, NA, vsoa_ASD),
        vsoa_diff = vsoa_ASD - vsoa_NA
    ) |>
    drop_na()
d_vsoa_diff_old <- readRDS("data/vsoa-autistic-nonautistic-diff-old.rds")
d_vsoa_diff_old_trimmed <- d_vsoa_diff_old |>
    mutate(
        vsoa_NA = if_else(vsoa_NA <   1,  1, vsoa_NA),
        vsoa_NA = if_else(vsoa_NA > 680, NA, vsoa_NA),
        vsoa_ASD = if_else(vsoa_ASD <   1,  1, vsoa_ASD),
        vsoa_ASD = if_else(vsoa_ASD > 680, NA, vsoa_ASD),
        vsoa_diff = vsoa_ASD - vsoa_NA
    ) |>
    drop_na()
d_ppt <- d_cdi |>
    select(subjectkey, nproduced, sex, group, interview_age) |>
    distinct()
d_aoa <- readRDS("data/CDI_English (American)_WS_14May2025_AoA.rds")
d_vsoa_aoa <- left_join(
    d_vsoa |> filter(group != "ASD-NA"),
    d_aoa |> select(num_item_id, aoa = aoa_glm), by = "num_item_id"
) |>
    mutate(
        group = factor(group, c("ASD", "NA"), c("autistic", "not aut."))
    ) |>
    relocate(aoa, .before = vsoa)
d_vsoa_aoa_trimmed <- d_vsoa_aoa |>
    mutate(
        vsoa = if_else(vsoa <   1,  1, vsoa),
        vsoa = if_else(vsoa > 680, NA, vsoa),
        aoa  = if_else( aoa <  12, 12,  aoa),
        aoa  = if_else( aoa >  30, NA,  aoa)
    ) |>
    drop_na()

```

## Participant Characteristics {.smaller}

```{r participant characteristics}
d_ppt |>
    mutate(
        group = factor(group, c("ASD", "NA"), c("autistic", "not aut."))
    ) |>
    group_by(group) |>
    summarize(
        across(
            c(nproduced, interview_age),
            list(mean = mean, median = median, sd = sd)
        )
    ) |> 
    kable(col.names = c("group", "mean", "median", "SD", "mean", "median", "SD"), digits = 3) |>
    kableExtra::add_header_above(header = c(" " = 1, "Productive Vocabulary" = 3, "Interview Age" = 3))

d_ppt |>
    mutate(
        group = factor(group, c("ASD", "NA"), c("autistic", "not aut.")),
        sex = factor(sex, c("M", "F"), c("male", "female"))
    ) |>
    count(group, sex) |>
    pivot_wider(id_cols = group, names_from = sex, values_from = n) |>
    mutate(total = male + female, `ratio (M:F)` = male / female) |>
    kable(digits = 3)
```

## VSOA correlation with AoA (raw) {.smaller}

Number of words contributing to correlations:

```{r number of words in correlation analysis}
d_vsoa_aoa |> drop_na() |> count(group) |> kable()
```

```{r vsoa correlation with aoa}
map(
    c("pearson" = "pearson", "spearman" = "spearman", "kendall" = "kendall"),
    function(df, method) {
        df |>
            group_by(group) |>
            summarize(r = cor(aoa, vsoa, method = method, use = "complete.obs"))
    },
    df = d_vsoa_aoa
) |>
    list_rbind(names_to = "method") |>
    pivot_wider(id_cols = group, names_from = method, values_from = r) |>
    kable(digits = 3)
ggplot(d_vsoa_aoa, aes(x = aoa, y = vsoa)) +
    geom_point() +
    geom_smooth(method = lm) +
    facet_wrap(~group) +
    theme_bw(base_size = 18)
```

## VSOA correlation with AoA (trimmed)  {.smaller}

```{r number of words in correlation analysis trimed}
d_vsoa_aoa_trimmed |> drop_na() |> count(group) |> kable()
```

```{r vsoa correlation with aoa trimmed}
map(
    c("pearson" = "pearson", "spearman" = "spearman", "kendall" = "kendall"),
    function(df, method) {
        df |>
            group_by(group) |>
            summarize(r = cor(aoa, vsoa, method = method, use = "complete.obs"))
    },
    df = d_vsoa_aoa_trimmed
) |>
    list_rbind(names_to = "method") |>
    pivot_wider(id_cols = group, names_from = method, values_from = r) |>
    kable(digits = 3)
ggplot(d_vsoa_aoa_trimmed, aes(x = aoa, y = vsoa)) +
    geom_point() +
    geom_smooth(method = lm) +
    facet_wrap(~group) +
    theme_bw(base_size = 18)
```

## How many words have extreme VSOA?

* Low: $\textrm{VSOA} \le 1$
* High: $\textrm{VSOA} \ge \max \textrm{vocab size}$

```{r extreme vsoa}
d_vsoa |>
    filter(group != "ASD-NA") |>
    mutate(
        extreme = if_else(vsoa <= 1, 1, if_else(vsoa >= max(d_ppt$nproduced), 2, NA)),
        extreme = factor(extreme, 1:2, c("low", "high"))
    ) |>
    drop_na(extreme) |>
    count(group, extreme) |>
    pivot_wider(id_cols = group, names_from = extreme, values_from = n) |>
    kable()
```

## Least-produced words (for autistic only; counts) {.scrollable}

```{r least produced words}
d_cdi_cumulative <- bind_rows(
    old = d_cdi |> filter(dataset == "old"),
    new = d_cdi,
    .id = "dataset_cumulative"
)

least_produced <- d_cdi_cumulative |>
    group_by(dataset_cumulative, group, word) |>
    summarize(n = sum(produced)) |>
    group_by(dataset_cumulative, group) |>
    slice_min(n, n=10)
    
least_produced_x <- d_cdi_cumulative |>
    filter(word %in% least_produced$word) |>
    group_by(dataset_cumulative, group, word) |>
    summarize(n = sum(produced), p = n / n()) |>
    mutate(
        least_10_asd = word %in%
            pull(filter(least_produced, group == "ASD", dataset_cumulative == "old"), word),
        least_10_non = word %in%
            pull(filter(least_produced, group == "NA", dataset_cumulative == "old"), word),
        least_10_for = if_else(
            least_10_asd & least_10_non,
            "both",
            if_else(
                least_10_asd & !least_10_non,
                "ASD",
                "NA"
            )
        )
    )

least_produced_counts <- least_produced_x |>
    pivot_wider(
        id_cols = c(least_10_for, word),
        names_from = c(dataset_cumulative, group),
        values_from = n,
        values_fill = 0
    ) |>
    select(-least_10_asd, -least_10_non) |>
    relocate(least_10_for, new_ASD, old_ASD, new_NA, old_NA, .after = word)

least_produced_proportions <- least_produced_x |>
    pivot_wider(
        id_cols = c(least_10_for, word),
        names_from = c(dataset_cumulative, group),
        values_from = p,
        values_fill = 0
    ) |>
    select(-least_10_asd, -least_10_non) |>
    relocate(least_10_for, new_ASD, old_ASD, new_NA, old_NA, .after = word)
```


```{r least produced words for ASD only counts}
least_produced_counts |>
    filter(least_10_for == "ASD") |>
    select(-least_10_for) |>
    kable(col.names = c("word", "Least 10 for", "New", "Old", "New", "Old"), digits = 2) |>
    kableExtra::add_header_above(header = c(" " = 2, "Autistic" = 2, "Not Autistic" = 2))
```


## Least-produced words (for autistic only; proportions) {.scrollable}

```{r least produced words p}
least_produced_proportions |>
    filter(least_10_for == "ASD") |>
    select(-least_10_for) |>
    relocate(new_ASD, old_ASD, new_NA, old_NA, .after = word) |>
    kable(col.names = c("word", "New", "Old", "New", "Old"), digits = 2) |>
    kableExtra::add_header_above(header = c(" " = 1, "Autistic" = 2, "Not Autistic" = 2))
```



## Least-produced words (for non-autistic only; counts) {.scrollable}

```{r least produced words for NA only counts}
least_produced_counts |>
    filter(least_10_for == "NA") |>
    select(-least_10_for) |>
    kable(col.names = c("word", "Least 10 for", "New", "Old", "New", "Old"), digits = 2) |>
    kableExtra::add_header_above(header = c(" " = 2, "Autistic" = 2, "Not Autistic" = 2))
```


## Least-produced words (for non-autistic only; proportions) {.scrollable}

```{r least produced words for NA only proportions}
least_produced_proportions |>
    filter(least_10_for == "NA") |>
    select(-least_10_for) |>
    relocate(new_ASD, old_ASD, new_NA, old_NA, .after = word) |>
    kable(col.names = c("word", "New", "Old", "New", "Old"), digits = 2) |>
    kableExtra::add_header_above(header = c(" " = 1, "Autistic" = 2, "Not Autistic" = 2))
```


## Words with significantly different VSOA between groups

### New VSOAs
```{r significant differences new}
d_vsoa_diff |>
    mutate(
        produced_earlier_by = if_else(sign(vsoa_diff) < 0, 1, 2) |>
            factor(1:2, c("autistic", "not aut.")),
        significant = sign(ci_u_diff) == sign(ci_l_diff)
    ) |>
    filter(significant) |>
    count(produced_earlier_by) |>
    kable()
```

### Old VSOAs
```{r significant differences old}
d_vsoa_diff_old |>
    mutate(
        produced_earlier_by = if_else(sign(vsoa_diff) < 0, 1, 2) |>
            factor(1:2, c("autistic", "not aut.")),
        significant = sign(ci_u_diff) == sign(ci_l_diff)
    ) |>
    filter(significant) |>
    count(produced_earlier_by) |>
    kable()
```

## Words with largest VSOA differences  {.smaller}

### Autistic children learn earlier
```{r autistic children learn earlier}
d_vsoa_diff |>
    select(-starts_with("ci_")) |>
    slice_min(vsoa_diff, n = 7) |>
    kable(digits = 1)
```


## Words with largest VSOA differences  {.smaller}

### Autistic children learn later
```{r autistic children learn later}
d_vsoa_diff |>
    select(-starts_with("ci_")) |>
    slice_max(vsoa_diff, n = 7) |>
    kable(digits = 1)
```


## Words with largest VSOA differences (trimmed)  {.smaller}

### Autistic children learn earlier
```{r autistic children learn earlier trimmed}
d_vsoa_diff_trimmed |>
    select(-starts_with("ci_")) |>
    slice_min(vsoa_diff, n = 7) |>
    kable(digits = 0)
```


## Words with largest VSOA differences (trimmed) {.smaller}

### Autistic children learn later
```{r autistic children learn later trimmed}
d_vsoa_diff_trimmed |>
    select(-starts_with("ci_")) |>
    slice_max(vsoa_diff, n = 7) |>
    kable(digits = 0)
```
